Profiles of Wind Direction and Speed in Turbulent Ekman Flow

### 2021-10-26 write1/manuscript.tex
continued writing on Section 4 -- ekman profile; spanwise profile; started on streamwise profile

\usepackage[margin=3.5cm]{geometry}
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{xcolor}
\usepackage{dirtytalk}
\usepackage{colortbl}
\usepackage{booktabs}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
At the small-scale end, the directional shear of the wind in the upper part of the surface layer
may cause a systmatic yaw for tall wind power generation devices where blades reach into the Ekman layer,
i.e. that part of the boundary layer where the wind starts to turn. 
%
\par
%
Wind veer in the planetary boundary layer is commonly characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a pre-requisite for any quantitative theory of the surface layer and a first
attempt in theoretical quantification was undertaken by \citet{rossby:PPO1935} using integral relations in the boundary layer.
%
Asymptotic similarity theory was later used by \citet{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}. 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\citet{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

@@

We denote the normalized modulus of the surface friction by

@@

These yield a Reynolds number as non-dimensional parameter governing the system.

@@

by new simulations at higher Reynolds numbers up to $1.28\times10^6$ with a horizontal domain extent up to $3.3\times 10^4$ viscous units. 

@@

The profiles of the shear-aligned streamwise velocity component are well-collapsed for the Reynolds
numbers investigated below $z^-\approx 0.15$ (circles in Fig.~\ref{fig:profiles}a); the logarithmic
law is appropriate for $50 < z^+ < 0.15 Re_\tau$, where $z^+ = Re_\tau z^-$.

@@

normalization. \todo{WHY? cf. TKE/Dissipation normalization above}

@@

agree well beyond $z^-\approx 0.3$ whereas the profiles of the streamwise component collapse also
within the overlap layer that extends much closer down to the surface

@@

which has indeed already been used by \cite{spalart:JFM1989}. 

@@

The normalized viscous stress

@@

The inner normalization of $\nu_E$ is obtained by division through the molecular viscosity, and it is $\nu_E^+=\nu_E/\nu = S_\mathrm{turb}/S_\mathrm{visc}$. Under this normalization, the profiles of eddy viscosity collapse below $z^+\approx 20$ with a tendency towards better collapse

@@

\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}

@@

While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to, understand this as the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.

@@

The classic text-book solution of Ekman flow makes use of the physical boundary conditions (BCs) for the atmospheric
boundary layer (no-slip at the bottom and geostrophic wind in the free atmosphere).

@@

These two boundary conditions eliminate one mode of the analytical solution and determine the magnitude of
the spiral.

@@

the dynamics put forth by \cite{ekman:AMA1905} are not appropriate in the surface layer of the ABL, where
the logarithmic, buffer, and viscous sublayers are found. 

@@

We to use a virtual boundary condition instead that takes into account the increased gradients in vicinity of
the surface and matches the magnitude of the Ekman spiral to the values observed in our DNS:

@@

@@

  G^{-1}\left(\begin{array}{c} 

@@

  + a_\mathrm{ek} e^{-z_\mathrm{ek}} \left(\begin{array}{r}

@@

  \end{array}\right)  \\
  z_\mathrm{ek} &=& \delta_{ek} (y^- -s_{ek}). 

@@

This solution is characterized by three parameters,

@@

(ii) the magnitude parameter of the spiral $a_{ek}$, and
(iii) a zero-crossing point for the velocity $s_{ek}$.
\begin{enumerate} 

@@

\item[(i)] The Ekman-layer depth scale $\delta_\mathrm{ek}$ is fundamentally defined by the eddy diffusivity.

@@

\item[(ii)] The magnitude parameter of the spiral primarily defines the super-geostrophic maximum of the

@@

    U_\mathrm{ek}/u_\star-Z_\star \propto a_\mathrm{ek} Z_\star = 8.4G.  \notag

@@

\item[(iii)] While $\delta_\mathrm{ek}$ sets the thickness across which the wind veering takes place,
  the zero-crossing height $s_\mathrm{ek}$ defines an offset.

@@

  From our DNS data, we estimate $s_\mathrm{ek}=-0.12\delta$. 
\end{enumerate} 
\todo{Paragraph and figure on idealized Ekman profile}

@@

The outer layer of Ekman flow is characterized by a turning of the wind velocity and the super-geostrophic maximum
that is sustained by momentum convergence at the inflection point of the velocity profile.

@@

  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
    \label{fig:profiles_outer} 

@@

In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
and the mean velocity is described by a viscous profile of the form
  (U,V)^{\alpha_\star+}  = (z^+,0)  
where the direction of the velocity points into the exact opposite direction of the wall shear stress $\tau$. 
In absence of roughness elements and for small roughness ($z_0^+<5$), this linear regime is known as
viscous sub-layer \cite{foken:m2002,foken:BM1978}.
In fact, this law of the wall has no degree of freedom given the drag, i.e. once $u_\star$ and $\alpha_\star$ are defined. 
However, theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
aloft.
A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
profile overestimate the velocity in the buffer layer. We therefore introduce a higher-order compensating term into
the linear profile before blending it into the logarithmic law:
  U_\text{inner}^{\alpha_\star}^+ = \frac{ y^+ + \gamma_4 (y^+)^{4} + \gamma_6 (y^+)^{6} } { 1 + \gamma_6 /u_{ref} (y^+ )^6}, 
where the fourth- and sixth-order terms in the nominator are designed to fit the profile below $y^+\approx 10$ and
the sixth-order correction in the denominator limits the growth of $u_\text{inner}$ for large $y^+$.

@@

@@

  \caption{Shear-aligned profiles of velocity components $U^{\alpha_\star+}$ in inner (left) and outer (right) units.} 

@@

velocity which challenges the conventional assumptions related to the so-called channel-flow analogy:

@@

While the universal profiles in vicinity of the wall implies that the profile be zero or at least small with respect to the streamwise
component, the veering requires a value of $V_{top}=U_G\sin\alpha_\star$ in the free stream (and thus also at the top of the boundary layer if we assume
that substantial velocity gradients are confined to the turbulent part of the flow). 

@@

This is commonly shown in terms of velocity hodographs aligned with the outer, geostrophic flow (cf. Fig.~\ref{fig:hodograph}) and normalized by

@@

Thus, the drag-law already implies three important properties of these hodographs,

@@

 \underline{U}^{G}(z=0) &=& 0,\\ 
 \underline{U}^{G}(z\mapsto\infty) &=& \underline{G}^G = \left(\begin{array}{c}1\\0\end{array}\right) \\
   \left.\frac{\partial V^G}{\partial U^G}\right|_{z=0} &=& \sin\alpha_\star

@@

Outer scaling of the velocity profile further implies that the velocity deficity of the spanwise component $V^{\alpha_\star}-G^{\alpha_\star}$
be a universal function of the outer height $z^-$, i.e.  
  V^{\alpha_\star}-G^{\alpha_\star} = f^{\alpha_\star}_V(z^-).  

@@

Interpreting the shear-aligned spanwise velocity deficit $f_{V}^{\alpha_\star}(z^-)$ as 
a signature of outer rotation may lead one to conclude that this universal function
is appropriate down to the surface--irrespective of the Reynolds number.

@@

This cannot be the case for the variation of $f_{V}^{\alpha_\star}(z^-)$ across the boundary layer (i.e. between $0<z^-<1$)
must fulfill the drag law and boundary conditions;
at the same time, the weak dependence of $\alpha_\star$ and $u_\star$ on $Re$ implies this range depends on $Re$.

@@

@@

\paragraph{Outer Layer.} In the outer region of the flow (for $z^-\mapsto 1$), $f_V(z^-)$, should govern the
spanwise velocity profile which indeed is supported by our DNS data (Fig.\ref{fig:outer_w}):

@@

The deficit is close to zero for $z^-\gtrsim 1$, and can be approximated by a logarithmic profile in the
range $0.15\lesssim z^- \lesssim 0.4$.

@@

The slope of this profile is best fit by $\kappa_V=0.836 (1 + 150 / Re_D) \kappa$, where the correction term for low Reynolds number
may be omitted for $Re_D \gg 1000$.

@@

Based on Fig.~\ref{fig:outer_w}, we pick the offset as
a constant fraction (0.906) of $V^\alpha_G=Z_\star \sin\alpha$, i.e
  V_\text{log}(z^-)  = \frac{1}{\kappa_V} \log\left(\frac{z^-}{0.4}\right) + 0.906V_G^{\alpha}. 

@@

@@

The eventually converged simulations at intermediate $Re_D$ (where the DNS could be run over several inertial periods)
exhibit a super-geostrophic maximum (positiv velocity deficit) in the span-wise component.

@@

This maximum is not present or less pronounced for the simulations at higher $Re$ where closer analysis reveals that the
velocity profiles in the outer layer are still subject to a small though finite adaptation process on the inertial time
scales $2\pi/f$ due to the relatively short time span for which these cases could be simulated.

@@

While there is no physical model for this wake region that leads to an analytical velocity profile formulation in the span-wise direction,
an error-function transition from the outer logarithmic law to the boundary condition provides a reasonable fit:
  V_\text{outer}(z^-) = (1-w_V)V_\text{log}(z^-) + w_V V_G^\alpha
with
w_V = \frac{\mathrm{erf}\left[\xi_\text{outer} \log(z/z_\text{ref})\right]  + 1}{2}
where $\xi_\text{outer}$ is a transition scale that defines the width of the wake region and $z_\text{ref}$ defines the height.
Based on the DNS data, we choose $\xi_\text{outer}=2$ and let $z_\text{ref}$ such that $V_\text{log}(z_\text{ref}) = 0.98V_G^\alpha$.
There is, however, a degree of ambiguity in the exact choice of these parameters due to the imperfect convergence of the
DNS data to the statistical equilibrium.
%%% PAR
\subsub{Inner Layer.} It is important to note that the logarithmic layer observed here for the spanwise velocity does not
conincide with the log-region of the streamwise velocity but is rather a 'continuation' through the outer layer.
This is consistent with the recent finding of a second log-region in the outer region of Ekman flow
(... et al. (incl. Sullivan), JAS 2019).
Below this region, the gradients in span-wise velocity are rather small and the span-wise velocity
monotonically approaches its surface boundary condition $V(z=0)=0$.
While the streamwise velocity follows a universal inner scaling that has acquired its universal, $Re$-independent shape for $Re_D> \mathcal{O}\left(10^3\right)$, the span-wise component that defines how the velocity vector veers when the surface is approached,
does not collapse in inner units, and there is, most importantly no sign of convergence even at the highest Reynolds numbers for
which simulations where carried out.
Even though the simplest assumption $V=0$ is reasonable for the lower part of the surface surface layer ($z^-<10^{-3}$), it does
not appropriately capture the profile in the rest of the surface layer:

@@

This illustrates that the higher-order (in terms of the channel-flow analogy where $V=0$) correction follows a mixed scaling when the surface is
approached which has been demonstrated for higher-order terms in other types of flow \citep{mellado:BM2016}. 

@@

%
\begin{figure}
  \includegraphics[width=1.0\textwidth]{../plot/w_profile.pdf}
  \caption{Profiles of shear-aligned span-wise velocity $(W^\alpha)^+$ versus inner and outer height.
  Dashed lines show DNS data, thick, opaque lines are from the semi-emprical theory developed above. }
  \label{fig:outer_w} 
\end{figure}
\end{subequations}

\section{Laminar Ekman solution with consideration of inner layer}

>>>>

\usepackage[margin=3.5cm]{geometry}
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[footnotesize]{caption}
\usepackage{xcolor}
\usepackage{dirtytalk}
\usepackage{colortbl}
\usepackage{booktabs}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
At the small-scale end, the directional shear of the wind in the upper part of the surface layer
may cause a systmatic yaw for tall wind power generation devices where blades reach into the Ekman layer,
i.e. that part of the boundary layer where the wind starts to turn. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a pre-requisite for any quantitative theory of the surface layer and a first
attempt in theoretical quantification was undertaken by \citet{rossby:PPO1935} using integral relations in the boundary layer.
%
Asymptotic similarity theory was later used by \citet{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}. 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\citet{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

@@

We denote the modulus of the surface shear, the surface friction, by

@@

In absence of external variability, this system converges to a statistically steady state in the sense that flow
statistics do not depend on time; and this state is defined by a Reynolds number,
the only non-dimensional parameter that governs the system.

@@

by new simulations at higher Reynolds numbers up to $\RE_\Lambda=1.28\times10^6$ with a horizontal domain extent up to $3.3\times 10^4$ viscous units. 

@@

The profiles of the shear-aligned streamwise velocity component are well-collapsed for $\RE_D>500$
below $z^-\approx 0.15$ (circles in Fig.~\ref{fig:profiles}a); the case with $\RE_D=500$ is only transitionally
turbulent and there is no distinct inner-oouter scale separation.
The logarithmic law is appropriate for $50 < z^+ < 0.15 Re_\tau$, where $z^+ = Re_\tau z^-$.

@@

normalization.
%\todo{WHY? cf. TKE/Dissipation normalization above}

@@

agree well beyond $z^-\approx 0.3$. %
This is a much higher level in comparison with the streamwise component that collapses
also within the overlap layer, i.e. much closer to the surface

@@

which has indeed already been found by \cite{spalart:JFM1989}, who estimate the constant from
an integral relation. 

@@

The viscous stress

@@

The inner normalization of $\nu_E$ is obtained when dividing by the molecular viscosity:
  \nu_E^+=\nu_E/\nu = S_\mathrm{turb}/S_\mathrm{visc}.
Under this normalization, the profiles of eddy viscosity collapse below $z^+\approx 20$ with a tendency towards better collapse

@@

\section{A universal velocity profile for the turbulent Ekman layer}

@@

Such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to.
This need for a connection of the two reference frames is a manifestation of a
mismatch in the theoretical treatment of the inner and outer layer in this rotating flow configuration. 

@@

The text-book solution for Ekman flow makes use of the physical boundary conditions (BCs) for the atmospheric
boundary layer (no-slip at the bottom and geostrophic wind in the free atmosphere) and a constant eddy
viscosity.

@@

Specifying the boundary conditions at top and bottom eliminates one mode of the analytical solution, and
it determines the magnitude of the spiral.

@@

The dynamics put forth by \citeauthor{ekman:AMA1905} in \citeyear{ekman:AMA1905}
are not appropriate in the surface layer of the ABL; better approximations exist for the logarithmic,
buffer, and viscous sublayers. 

@@

In view of this situation, we use an adapted Ekman spiral that does not enforce the
boundary conditions at the surface but at a different height. 
This is achieved by considering the Ekman spiral only in the Ekman layer,
thus giving way for the well-established logarithmic and viscous-layer profiles in the
lower surface layer. 
Based on the derivation in App.~\ref{app:ekman_solution}, this profile is given by 

@@

  \label{eqn:profile_ekman}

@@

    \frac{1}{G}\left(\begin{array}{c} 

@@

    + e^{-z_\mathrm{ek}} \left[ a_\mathrm{ek}  \left(\begin{array}{r}

@@

      \end{array}\right)  + b_\mathrm{ek}\left(\begin{array}{r}
        \sin z_\mathrm{ek} \\ \cos z_\mathrm{ek}
        \end{array}\right)\right]\\
    z_\mathrm{ek} &=& \delta_\mathrm{ek} (y^- -s_\mathrm{ek}). 

@@

The right-hand-side consists of two modes with magnitude $a_\mathrm{ek}$ and $b_\mathrm{ek}$
shifted by $\pi/2$ with respect to each other. 
In the classic case, the second mode governed by $b_\mathrm{ek}$ is incompatible with
the surface boundary condition. 
While this is not the case here for the general form of the profile, we
phase shifts can also be captured by the parameter $s_\mathrm{ek}$, 
and we stick with to a single-modal approach, i.e., we let $b_\mathrm{ek}=0$. 
This single-modal solution is characterized by three parameters,

@@

(ii) the magnitude parameter of the spiral $a_\mathrm{ek}$, and
(iii) a zero-crossing point for the velocity $s_\mathrm{ek}$.

@@

The effects of varying these parameters are illustrated in Fig.~\ref{fig:ekman_ideal} where
the classic Ekman solution is recovered by setting $a_\mathrm{ek}=1$, $s_\mathrm{ek}=0$ and
$\delta_\mathrm{ek}=\sqrt{2\nu/f}$.
These parameters are \emph{a priori} unknown as they need to conform to the turbulent state of the
boundary layer; we use our DNS data to arrive at best estimates for them.
    \textbf{\ \ a)}\hspace{0.45\textwidth}\textbf{\ \ b)}\\
  \includegraphics[width=\textwidth]{../plot/ekman_ideal.pdf}
  \caption{\label{fig:ekman_ideal}
    \textbf{a)} Generalized Ekman-profile of the geostrophic-aligned component $U_\mathrm{ek}$.
    \textbf{b)} Hodograph for the geostrophic-aligned and pressure-gradient aligned components $U_\mathrm{ek}$ and $V_\mathrm{ek}$.
    Thick, black dashed line shows the classic solution. The height corresponding to $z^-=0.30$ is marked by the dashed line in panel \textbf{(a)}
    and by filled circles in panel \textbf{(b)}. The hodograph and profiles above this reference height are
    shown as solid lines, below as opaque line.
\textbf{The Ekman-layer depth scale} $\delta_\mathrm{ek}$ is fundamentally defined by the eddy viscosity.

@@

\textbf{The magnitude parameter of the Ekman spiral}, $a_\mathrm{ek}$, defines the super-geostrophic maximum of the

@@

  U_\mathrm{ek}/u_\star-Z_\star \propto a_\mathrm{ek} Z_\star = 8.4G.
  \label{eqn:hodograph_scaling}

@@

Thus, the reduction of the area under the Ekman hodograph is directly linked to the variation
in the friction velocity, which quantifies the well-known qualitative observation that a more turbulent boundary
layer leads to a flatter hodograph in Ekman flow.
\textbf{The offset parameter} $s_\mathrm{ek}$ defines the zero-crossing height of the profile
(in contrast to $\delta_\mathrm{ek}$, which determines the thickness across which the wind veering takes place). 

@@

From our DNS data, we estimate $s_\mathrm{ek}=-0.12$. 

@@

In summary, the outer layer of Ekman flow is characterized by a turning of the wind
velocity and the super-geostrophic maximum that is sustained by momentum convergence
at the inflection point of the velocity profile.

@@

Corresponding profiles are shown in comparison with data from three DNS runs in Fig.~\ref{fig:profiles_outer}.
The idealized profiles capture the secondary minimum and convergence to the geostrophic equlibrium in the
non-turbulent flow very well.

\begin{figure}
  \begin{flushleft}
    \textbf{\ \ a)}\hspace{0.45\textwidth}\textbf{\ \ b)}\\
  \end{flushleft} 
  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
  \caption{
    \label{fig:profiles_outer}
    Shear-aligned velocity deficit for the wall-streamwise (panel \textbf{(a)}) and wall-spanwise (panel \textbf{(b)})
    components of the mean velocity $U^\alpha$ and $V^\alpha$.
    Solid lines show DNS data, dashed lines the Ekman profiles $U_\mathrm{ek}$ and $V_\mathrm{ek}$ as defined in
    Eq.~\ref{eqn:profile_ekman}. Region below $z^-=0.3$ is shown opaque. Variations in $U_\mathrm{ek}$ and $V_\mathrm{ek}$ are
    a consequence of the normalization and related to chagnes in $u_\star$ and $\alpha_\star$ among the different $\RE_D$. 
  }
\end{figure}

@@

In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
and the mean velocity is described by a viscous profile of the form
\begin{equation}
  U^{\alpha_\star+}  = z^+
\end{equation}
where the direction of the velocity points into the exact opposite direction of the wall shear stress $\tau$. 
In absence of roughness elements and for small roughness ($z_0^+<5$), this linear regime is known as
viscous sub-layer \cite{foken:m2002,foken:BM1978}.
%
In fact, this law of the wall has no degree of freedom given the drag, i.e. once $u_\star$ and $\alpha_\star$ are defined. 
%
However, theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
aloft.
A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
profile overestimate the velocity in the buffer layer. We therefore introduce a higher-order compensating term into
the linear profile before blending it into the logarithmic law:
\begin{equation}
  U_\text{inner}^{\alpha_\star}^+ = f_{U,\mathrm{inner}}(z^+) = \frac{ z^+ + \gamma_4 (z^+)^{4} + \gamma_6 (z^+)^{6} } { 1 + \gamma_6 /u_{ref} (z^+ )^6}, 
\end{equation} 
where the fourth- and sixth-order terms in the nominator are designed to fit the profile below $y^+\approx 10$ and
the sixth-order correction in the denominator limits the growth of $u_\text{inner}$ for large $y^+$.
%
These correction terms are small for $z^+<5$ where the linear law of the wall
(shown in the inset in Fig.~\ref{fig:u_profiles}) yields an excellent fit of the data; we use here
\begin{equation*}
    \gamma_4 = -3.825 \times 10^{-4};\quad
    \gamma_6 &=&  6.320 \times 10^{-6};\quad 
    u_\mathrm{ref} &=& 1.278 \times 10^{1}.
\end{equation*}

@@

    \textbf{\ \ a)}\hspace{0.45\textwidth} \textbf{\ \ b)} 

@@

  \caption{Shear-aligned profiles of velocity components $U^{\alpha_\star+}$ in inner (left) and outer (right) units.
    \label{fig:u_profiles}} 

@@

velocity which challenges the conventional assumptions related to the channel-flow analogy:

@@

While the analogy with channel flow in vicinity of the wall implies that the streamwise component
be zero or at least small, the veering requires a value of $V_{top}=U_G\sin\alpha_\star$ in the free stream
(and thus also at the top of the boundary layer if we assume that any substantial velocity gradient is
confined to the turbulent part of the flow). 

@@

This continuous rotation of the wind vector is conveniently visualized by velocity hodographs
aligned with the outer, geostrophic flow (cf. Fig.~\ref{fig:ekman_ideal}b) and normalized by

@@

The geometry of the flow and its drag imply the following for any hodograph: 

@@

  V^{\alpha_\star}(z=0) &=& 0,\\ 
  \lim_{z\rightarrow\infty} V^{\alpha_\star} &=& G \sin\alpha_\star \\ 
  \left.\partial_{z^+} V^{\alpha_\star+}}\right|_{z=0} &=& 0.  

@@

@@

Outer scaling of the velocity profile further implies that the velocity deficit of
$(V^{\alpha_\star}-G^{\alpha_\star})/u_\star$ be a universal function of the outer height $z^-$. 

@@

In the outer region of the flow (for $z^-\mapsto 1$), $f_V(z^-)$, should govern the
spanwise velocity profile, as is supported by our DNS data (Fig.\ref{fig:profiles}b);
above $z^-\approx 0.3$, this profile is very well approximated by the Ekman-turning derived
above (Eq.~(\ref{eqn:profile_ekman}); Fig.~\ref{fig:profiles_outer}b). 

@@

While this deficit is a signature of outer rotation,
it is inappropriate to extend this general relation to the surface where inner
scales matter:

@@

On the one hand, the variation of the spanwise velocity deficit across
the boundary layer (i.e. between $0<z^-<1$) must match the difference implied by the drag law ($u_\star, \alpha_\star$)
and the constant value of $V^{\alpha_\star}$ around $z^-=0.3$.

@@

On the other hand,--provided the outer velocity deficit is $\RE$ independent--the $\RE$-dependence
of $\alpha_\star$ and $u_\star$ implies that this difference cannot be constant as a function of $\RE$

@@

@@

We hence ask, how does the span-wise component scale when the surface is approached?

@@

Clearly, the spanwise contribution is small in comparison with the streamwise component
throughout much of the layer below $z^-\approx 0.3$. 

@@

However, we cannot assume $V=0$ if a smooth matching between the inner and outer layers shall be achieved. 

@@

In this context, we first realize that the velocity deficit $(V^{\alpha_\star}-G^{\alpha_\star})/u_\star$
approaches a $\RE$-independent constant around $C_{V0}=Z_\star \sin\alpha -6.1$ at the surface; deviations from
this constant are only found for the lowest Reynolds numbers which is in accordance with
the low-$\RE$ correection suggested by \cite{spalart:JFM1989}.

@@

This constrains the wind veer, and it quantitatively shows that the decreasing wall friction manifest in an
increase of $Z_\star$ exactly compensates the decrease of wind turning measured by $\sin\alpha_\star$.

If the difference across the boundary layer is constant ($C_{V0}$) across variations of $\RE$,
the averaged gradient $\partial_{z^+} V^{\alpha_\star+}$ of the spanwise velocity component
must decrease as $1/\delta^+$ as $Re_\tau$ increases.
%
Hence, it should--at a fixed height--be $V^{\alpha_\star} \propto \left(\delta^{+}\right)^{-1}$.
%
A profile that agrees with the constraints of the profile at the surface
and exploits the dependence of $V^{\alpha_\star}$ on $\delta^+$ is 
\begin{equation}
  V^{\alpha_\star}\frac{\delta^+}{G} = f_{V,\mathrm{inner}}(z^+) = v_\mathrm{ref} \left( \omega_v z^+ -1 + \exp[-\omega_v z^+]\right),
  \label{eqn:scaling_spanwise} 
\end{equation}
where $v_\mathrm{ref}$ controls the magnitude of the profile and $\omega_v$ sets the height at which the profile
transitions into an approximately linear one.
%
We find excellent agreement with the DNS data for $500\le\RE_D\le1600$ below $z^+\approx 15$ with
\[v_\mathrm{ref} = 18.85;  \quad \omega_v=0.2353\]
(cf. Fig.~\ref{fig:inner_w}b). 
%
\par
%
Eq.~(\ref{eqn:scaling_spanwise}) establishes a universal mixed scaling for the spanwise velocity:
%
While it requires the vertical coordinate to be expressed in inner units,
the velocity itself is normalized by the geostrophic wind, and becomes
inversely proportional to the friction Reynolds number $\RE_\tau=\delta^+$ when considered at a fixed height.
%
In vicinity of the surface, such mixed scaling has already been identified for higher-order statistics in convective 
flows \citep{mellado:BM2016,li:JAS2018}, where large scales leave their signatures in vicinity of the surface.
%
It is important to note here that, while $V$ is a first-order statistic from a statistical perspective,
the spanwise velocity is a higher-order correction term from the view point of similarity theory and
the channel-flow analogy that is routinely employed in the surface layer.
%
Further, this is consistent with the scaling for the velocity hodograph found in Eq.~(\ref{eqn:hodograph_scaling}) where
the friction velocity also drops out.
%
\par
%
The universal scaling of $V^{\alpha_\star}/G\delta^+$ continues through the surface layer,
and we find a very similar picture to the spanwise component with an increasing
depth of a quasi-logarithmic layer as shown in Fig.~\ref{fig:inner_w}b.
%
We model this layer by a profile of the form
%
\begin{equation}
  f_{V,\mathrm{log}}(z^+) = \frac{V_\mathrm{log}(z^+)}{G} \delta^+ = a_\mathrm{log} + b_\mathrm{log} \log z^+ + c_\mathrm{log} z^+
\end{equation} 
%
This profile is constrained by
\begin{align}
  f_{V,\mathrm{log}}(z^+=10) &=& f_{V,\mathrm{inner}}(z^+=10) &=:& v_{10} \simeq 27.3\\
  \frac{\p}{\p z^+} \left[ f_{V,\mathrm{log}}\right]_{z^+=10} &=& \frac{\p}{\p z^+}\left[ F_{V,\mathrm{inner}}\right]_{z^+=10} &=:&d_{10}\simeq 4.01 \\
  f_{V,\mathrm{log}}(z^+=0.3\delta^+) &=& V^{\alpha_\star}_\mathrm{ek}(z^-=0.3)\delta^+&=:& v_{03} 
\end{align}
where $v_{03}$ is determined by $V_\mathrm{ek}(0.3)$ and $U_\mathrm{ek}(0.3)$ and depends on the Reynolds number.

%
\begin{figure}
  \begin{flushleft}
    \textbf{\ \ a}\hspace{0.45\textwidth}\textbf{\ \ b}\\
  \end{flushleft} 
  \includegraphics[width=1.0\textwidth]{../plot/w_profile.pdf}
  \caption{Profiles of shear-aligned span-wise velocity $(W^\alpha)^+$ versus inner height. 
    Dashed lines show DNS data, thick, opaque lines are from the semi-emprical theory developed above.
    Left panel show standard inner normalization. Right panel shows the inviscid normalization yielding
    a universal profile for the spanwise component of velocity in the inner layer. }
  \label{fig:inner_w}
\end{figure}
\end{subequations}

\section{Laminar Ekman solution with consideration of inner layer}
\label{app:ekman_solution}

\appendix
\section{Old Stuff}
\todo{OLD FROM HERE} 
Below this region, the gradients in span-wise velocity are rather small and the span-wise velocity
monotonically approaches its surface boundary condition $V(z=0)=0$.
%
While the streamwise velocity follows a universal inner scaling that has acquired its universal, $Re$-independent shape for $Re_D> \mathcal{O}\left(10^3\right)$, the span-wise component that defines how the velocity vector veers when the surface is approached,
does not collapse in inner units, and there is, most importantly no sign of convergence even at the highest Reynolds numbers for
which simulations where carried out.
%
Even though the simplest assumption $V=0$ is reasonable for the lower part of the surface surface layer ($z^-<10^{-3}$), it does
not appropriately capture the profile in the rest of the surface layer:

@@

@@
